\paperTitle

### 2014-05-17 doc/icml2014/directed.tex
Fixed typos in piecewise

It is clear that having uniform bottlenecks implies part (ii)
of the exclusive views property (\propertyref{exclusive-views}) via \TensorFactorize,
but we will show that part (i) follows as well.

The exclusive views property 
Next, we define a (in analogy with biconnected component)
\begin{definition}[Bidependent set]
Given a graphical model $\sG$, we say that a subset of nodes $S$ is \emph{bidependent} if
conditioned on any $a \in S$, there is an active trail between any other two nodes $b,c \in S$.
\end{definition}
Intuitively
Note that all cliques are bidependent, but 
Bottlenecks provide three views, but to satisfy condition (ii) of \propertyref{exclusive-views},
only two is required.

Now we establish the link between bottlenecks and exclusive views:
\begin{lemma}[Uniformly bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
% PL: simplify to cliques (this is also bad notation, since \sG is a set of cliques, not nodes)
%Let $\sG' = \{ h_{i_1}, \dots, h_{i_m} \} \subseteq \sG$ be
  %a connected subgraph of hidden variables.
  %If every hidden variable
  %$h_i \in \sG'$ is a bottleneck, then $\sG'$ has the exclusive views property.
  If a graphical model $\sG$ is faithful and uniformly bottlenecked, then
  every bidependent set in $\sG$ has the exclusive views property (and hence so does $\sG$).
\end{lemma}
\begin{proof}
We prove the contrapositive: suppose there exists a biconnected subgraph of $\sH \subset \sG$ consisting of hidden nodes
such that some $h_0 \in \sH$ does not have an exclusive view.
This means each observed variable $x_i$ in the graph is conditionally dependent on some
$h_i \in \sH \backslash \{h_0\}$ given $h_0$.
Consider any such two $x_1, x_2$.
Conditioned on $h_0$,
there is an active trail between $h_1$ and $h_2$ because $\sH$ is biconnected.
Thus, there is an active trail $x_1 - h_1 - h_2 - x_2$ conditioned on $h_0$.
Since $\sG$ is faithful, we have $x_1 \not\perp x_2 \mid h_0$.
As this holds for any $x_1,x_2$, $h_0$ cannot be a bottleneck.
\end{proof}

@@

>>>>

%It is clear that having uniform bottlenecks implies part (ii)
%of the exclusive views property (\propertyref{exclusive-views}) via \TensorFactorize,
%but we will show that part (i) follows as well.

First, define the notion of a bidependent set (in analogy with biconnected component),
in which conditionining on one variable does not make anything independent:
\begin{definition}[Bidependent set]
Given a graphical model $\sG$, we say that a subset of nodes $S$ is \emph{bidependent} if
conditioned on any $a \in S$, there is an active trail between any other two nodes $b,c \in S$.
\end{definition}
Note that all cliques are bidependent, but bidependent can have fewer dependencies.

Bottlenecks provide three views, but to satisfy condition (ii) of \propertyref{exclusive-views},
only two is required.

Now we establish the link between bottlenecks and exclusive views:
\begin{lemma}[Bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
  If a bidependent subset of hidden variables $S$ is a bottleneck,
  then $S$ has the exclusive views property.
\end{lemma}
\begin{proof}
We prove the contrapositive: suppose there exists a biconnected subgraph of $\sH \subset \sG$ consisting of hidden nodes
such that some $h_0 \in \sH$ does not have an exclusive view.
This means each observed variable $x_i$ in the graph is conditionally dependent on some
$h_i \in \sH \backslash \{h_0\}$ given $h_0$.
Consider any such two $x_1, x_2$.
Conditioned on $h_0$,
there is an active trail between $h_1$ and $h_2$ because $\sH$ is biconnected.
Thus, there is an active trail $x_1 - h_1 - h_2 - x_2$ conditioned on $h_0$.
Since $\sG$ is faithful, we have $x_1 \not\perp x_2 \mid h_0$.
As this holds for any $x_1$ and $x_2$, $h_0$ cannot be a bottleneck.
\end{proof}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \subsection{Sample complexity}
% \label{sec:sampleComplexity}
% PL: "combine" is too vague to be useful
%\LearnMarginals combines two consistent algorithms, \TensorFactorize and
%\LearnCliqueNs, and is thus consistent itself.

%In this section, we provide formal conditions under which $\LearnMarginals$
%will produce consistent estimates.
%We let
%$\mOpphat{v}{i}$,
%$\hat Z_\sC$,
%and $\hat M_\sV$,
%denote estimators for
%$\mOpp{v}{i}$,
%$Z_\sC$,
%and $M_\sV$,
%respectively.

% But $\mOpp{v}{i}$ is a product of tensors on the path from $h_i$ to
%   $x_v$, marginalizing out the rest of the graph.
% To make this property more interpretable, let us assume the following
%   property which we prove to be sufficient in
%   \appendixref{assumption-proof}:
% \begin{assumption} 
% \label{asm:full-rank-plus}
% The marginal distribution of every latent variable has full support:
%   $\forall{h_i \in H},~ \mPi{i} \succ 0$.
% For every clique $\sC \in \sG$ (including ones with observed variables),
%   and every conditional distribution $\Pr(h_c \given \Pa(h_c))$, where
%   $h_c \in \sC$, every {\em tensor unfolding} has full column rank, $k$.
% \end{assumption}
% In our directed grid example (\sectionref{directedExample}), for the
%   clique $\sC = \{ h_1, h_2 \}$ this condition implies $\pi = \Pr(h_1)
%   \succ 0$ and $T = \Pr(h_2 \mid h_1)$ has column rank $k$; for the clique
%   $\sC = \{ h_1, x_1 \}$, it implies $O = \Pr(x_1 \mid h_1)$ has column
%   rank $k$.

%\todo{did they prove this; if so just say "they show", not "we can show"}
%Using results from
%  \citet{anandkumar12moments,anandkumar13tensor} we can show that
%  learning $\mOpp{v_1}{i}$ for the bottleneck $h_i$ with views $x_{v_1},
%  x_{v_2}, x_{v_3}$ has the following sample complexity,
%  \todo{define hat O}
%  \todo{max subscript is misaligned}
%\begin{align*}
%  \|\mOpphat{v_1}{i} - \mOpp{v_1}{i}\|^2_F &= \frac{1}{\sqrt{n}} O\left( \frac{k {\pi\oft{i}}_{\max}^2}{\sigma_k(M_{v_1,v_2})^5} \right). 
%\end{align*}
%It is easy to see that $\LearnMarginals$ has a polynomial sample complexity because it composes two parts that individually have polynomial sample complexity.
%From \citet{anandkumar13tensor} given the non-degeneracy assumptions
%  \assumptionref{non-degeneracy}, we have that $\mOpphat{v}{i}$
%  converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant
%  that depends polynomially on the $k$-th singular value of
%  $\mOpp{v}{i}$.
% PL: we don't have a crisp theorem (because there's multiple paths), so leave it vague.
%Recovering the marginals $Z_\sC$ from the conditional moments via $\LearnClique$ is a linear operation and thus also has polynomial sample complexity. 
%Note that $\sigma_{k}(\mOpp{v}{i})$ can become extremely
%small if $h_i$ and $x_v$ are connected via many intermediate hidden variables
%\footnote{To see this, consider a chain-structured graphical model: $h_1
%\to h_2 \cdots \to h_t \to x_v$. In this example, if
%$\sigma_k(\Pr(h_{i+1} \given h_{i})) \ge \sigma_k$ for each $i = 1,
%\cdots, t-1$, then $\sigma_k(\mOpp{v}{1})$ can be as bad as
%$\sigma_k^t \sigma_k(\mOpp{v}{t})$.}.
%\begin{theorem}[Sample complexity for $\mOpp{v}{i}$]
%  \label{thm:sample-complexity}
%  Without loss of generality, let $v=1, i=1$, and let $x_1, x_2, x_3$ be three views for
%  $h_1$. With probability at least $1 - \delta$,
%\begin{align*}
%  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
%    &\le  \\
%    &
%  \hspace{-0.3in}
%      O\left(
%      % PL: remove pi_max because that's upper bounded by 1.
%      \frac{k \log(1/\delta) } 
%      {\sqrt{n} \, \pi\oft{1}_{\min} (\sigma_{k}(\mOpp{1}{1}) \sigma_{k}(\mOpp{1}{2}))^{5/2}} \right),
%\end{align*}
%where $\pi\oft{1}_{\min}$ is the smallest element of $\pi\oft{1}$.

%If $\sigma_{\min}(\Pr(c \given \Pa(h_c))) \le \sigma$ for every clique
%  $\sC \in \sG$ and hidden variable $h_c \in \sC$,
%\begin{align*}
%  \|\mOpphat{1}{1} - \mOpp{1}{1}\|_F 
%    &\le 
%      O\left( 
%      \frac{k \log(1/\delta) {\pi\oft{1}}_{\max}/{\pi\oft{1}}_{\min}} 
%      {\sqrt{n} \sigma^{5t}} \right) \epsilon,
%\end{align*}
%where $t$ is the length of the shortest topological ordering from $h_1$
%to the unique parent of $x_1$, $h_t$.
%\end{theorem}
%In general, $\sigma_{k}(\mOpp{v}{i})$ will depend exponentially on the
  %path length between the hidden variable $h_i$ and $x_v$.
%This is expected because the influence from $h_i$ to $x_v$
  %gets diluted for every latent variable it must factor through.

%\theoremref{sample-complexity}, proved in \appendixref{assumption-proof}
%  says that the number of samples required to estimate the view
%  $\mOpp{v}{i}$ grows exponentially in the {\em distance} between $x_v$
%  and $h_i$.

%\todo{Present sample complexity bounds for $Z_\sC$?}

% Easy case
% Next, we need to show that the hidden marginals $\hat Z_\sC$ converge to $Z_\sC$.
% First, $\hat M_\sV$ is just an empirical average of multinomials over the data points.
% Abusing notation slightly, we let $\hat M_\sV$ also denote its
% $d^{m}$-dimensional vectorized version.
% We also represent $\mH_\sC$ as
%   a vector in $\Re^{k^m}$, and represent $\mOppAll \eqdef
%   \mOpp{v_1}{i_1} \otimes \cdots \otimes
%   \mOpp{v_m}{i_m}$ as a matrix in $\Re^{d^m \times
%   k^m}$.
% By the central limit theorem, we have:
% $\sqrt{n} (\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV)$,
% where $\Sigma_\sV$ is the \emph{asymptotic variance} of $\hat M_\sV$: 
% % ARUN: I don't think so - \todo{doesn't this need to be multiplied by something?}
% \begin{align*}
% \Sigma_\sV \eqdef \dM_\sV - M_\sV M_\sV^\top, \quad \dM_\sV \eqdef \text{diag}(M_\sV).
% \end{align*}
% 
% Our next step will be to use the delta-method to convert the above
% result into the asymptotic variance for the pseudoinverse version of
% $\LearnClique$. 
% \begin{lemma}[Asymptotic variance of pseudoinverse estimator for $\tilde Z_\sC$]
%   \label{lem:mom-variance}  
%   Assume $\hat M_{\sV}$ has asymptotic variance $\Sigma_\sV$ defined above.
%   Then the asymptotic variance of $\hat{Z_\sC}$ is:
%   \begin{align*}
%     \Sigma^{\mom} &= \mOppAlli \Sigma_\sV \mOppAllit.
%   \end{align*}
%   %where ${\mOppAll}\pinv \eqdef {\mOpp{v_1}{i_1}}\pinv \otimes
%   %\cdots \otimes {\mOpp{v_m}{i_m}}\pinv$, a $d^m \times k^m$ matrix.
% \end{lemma}
% \begin{proof}
% For clique $\sC$, recall we have
%   $Z_{\sC} = \mO_\sV(\mOppi{v_1}{i_1},\cdots,\mOppi{v_m}{i_m})$.
% %Choosing to represent $Z_\sC$ and $M_\sV$ as vectors and
% %$\mOppi{v_1}{i_1} \otimes \cdots \otimes \mOppi{v_m}{i_m}$ as the
% %matrix,
% We can rewrite this representation more compactly as: $Z_{\sC} = {\mOppAll}\pinv \mO_\sV$.
% By the delta-method \cite{vaart98asymptotic},
% % DONE: need to shorten
% %we have that the asymptotic variance of $M_\sV$ is:
% %\begin{align*}
% %  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
% %\end{align*}
% %where $\Sigma_\sV$ is the variance of the observations. 
% we immediately get:
% $\sqrt{n}(\hat Z_\sC - Z_\sC) \convind \sN(0, \mOppAlli \Sigma_\sV \mOppAllit)$.
% \end{proof}

% DONE: too brazen to say
%Note that extending these results to finite sample bounds can be done
  %via a straightforward application of perturbation bounds.
